\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-23 doc/paper-jss/sdmTMB-paper.Rnw
Edits

% \widowpenalty10000
% \clubpenalty10000

@@

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
notation, and example code. An \pkg{sdmTMB} model can be built from any
combination of the process components (first six rows) plus an observation
component (last row). The examples are from an SDM context, but the model can be
fit to any spatially referenced point data. Notation: We refer to design
matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
symbols represent standard deviations and covariance matrices, respectively. All
other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
$\bm{\omega}$ refer to a vector of main effects and spatial random field
deviations, respectively). See the \pkg{sdmTMB}
\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
for a full description of the model. Note that \code{s()} denotes a smoother as
in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). 
The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()}. 
\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} and \INLA's triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}).
The formula syntax follows a standard \R\ format and is expanded on in the next paragraph.
A fitted model summary can be viewed with \texttt{print.sdmTMB()} or \texttt{summary.sdmTMB()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
Prediction on fitted or new data is accomplished with \texttt{predict.sdmTMB()}.
\texttt{residuals.sdmTMB()} calculates various types of residuals.
New observations can be simulated from a fitted model with \texttt{simulate.sdmTMB()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
\texttt{sdmTMB\_cv()} facilitates cross validation.
Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}). 
% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

@@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5.75>>=
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |>
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
@

@@

@@

<<mesh-vis-timing, fig.width=7, fig.asp=1.1, out.width="4in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
g <- list()
for (i in seq_along(max_edge_vec)) {
  .mesh <- INLA::inla.mesh.2d(
    boundary = segm_bnd,
    max.edge = c(max_edge_vec[i], 0.2),
    offset = c(0.1, 0.05)
  )
  g[[i]] <- ggplot() +
    gg(.mesh) +
    theme_light() +
    coord_equal(expand = FALSE) +
    ggtitle(paste("Mesh vertices =", .mesh$n))
}
cowplot::plot_grid(plotlist = g, ncol = 2)
@

>>>>

% \widowpenalty1000
% \clubpenalty1000

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
notation, and example code. An \pkg{sdmTMB} model can be built from any
combination of the process components (first six rows) plus an observation
component (last row). The examples are from an SDM context, but the model can be
fit to any spatially referenced point data. Notation: We refer to design
matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
symbols represent standard deviations and covariance matrices, respectively. All
other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
$\bm{\omega}$ refer to a vector of main effects and spatial random field
deviations, respectively). See the \pkg{sdmTMB}
\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
for a full description of the model. Note that \code{s()} denotes a smoother as
in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()}.
\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} and \INLA's triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.
Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Families are specified with standard R family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families for family's not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).
The formula syntax follows a standard \R\ format and is expanded on in the next paragraph.
A fitted model summary can be viewed with \texttt{print.sdmTMB()} or \texttt{summary.sdmTMB()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.
Prediction on fitted or new data is accomplished with \texttt{predict.sdmTMB()}.
\texttt{residuals.sdmTMB()} calculates various types of residuals.
New observations can be simulated from a fitted model with \texttt{simulate.sdmTMB()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
\texttt{sdmTMB\_cv()} facilitates cross validation.
Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

@@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5, fig.pos="htbp", out.width="4in">>=
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |>
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
@

@@

% \clearpage

@@

<<mesh-vis-timing, fig.width=7, fig.asp=1.1, out.width="4.3in", echo=FALSE, fig.cap="The meshes used in simulations from least to most vertices.", fig.pos='ht', fig.align='center'>>=
max_edge_vec <- rev(c(0.06, 0.075, 0.1, 0.15, 0.2))
g <- list()
for (i in seq_along(max_edge_vec)) {
  .mesh <- INLA::inla.mesh.2d(
    boundary = segm_bnd,
    max.edge = c(max_edge_vec[i], 0.2),
    offset = c(0.1, 0.05)
  )
  g[[i]] <- ggplot() +
    gg(.mesh) +
    theme_light() +
    coord_equal(expand = FALSE) +
    ggtitle(paste("Mesh vertices =", .mesh$n))
}
cowplot::plot_grid(plotlist = g, ncol = 2)
@
